---
title: "Lab on Discrete Choice"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: false
      smooth_scroll: false
---

 1. consider estimation of demand model for transport
 2. look at IIA assumption (potential pitt-falls)
 3. extend with 3 types, check substitution patterns
 4. consider max-rank substitution

We consider here discrete choices over a set of alernatives. THe utility of the agent is modeled as 

# Discrete Choices

$$u_i(j) = \beta X_i + \epsilon_{ij}$$

and when the error term is type 2 extreme value:

$$F(\epsilon_{ij}) = \exp( -\exp( -\epsilon_{ij}  ) )$$

the choice probability is given by 

$$ Pr[j(i)^*=j] = \frac{ \exp[ u_i(j)  ] }{ \sum_j'  \exp[ u_i(j')  ]} $$

Armed with these tools we can tackle the data we are given.

## Data
```{r,message=FALSE}
library(AER)
library(mlogit)
library(kableExtra)
library(knitr)
library(foreach)
data("TravelMode")
data = TravelMode

kable(data[1:10,])
```

## Mlogit results

```{r,message=FALSE}
library(AER)
library(mlogit)
library(kableExtra)
library(knitr)
library(reshape2)
data("TravelMode")
TravelMode <- mlogit.data(TravelMode, choice = "choice", shape = "long", alt.var = "mode", chid.var = "individual")
data = TravelMode

## overall proportions for chosen mode
with(data, prop.table(table(mode[choice == TRUE])))

## travel vs. waiting time for different travel modes
ggplot(data,aes(x=wait, y=travel)) + geom_point() + facet_wrap(~mode)

## Greene (2003), Table 21.11, conditional logit model
fit1 <- mlogit(choice ~ gcost + wait, data = data, reflevel = "car")
# fit1 <- mlogit(choice ~ gcost + wait | income, data = data, reflevel = "car")
# fit1 <- mlogit(choice ~ gcost + wait + income, data = data, reflevel = "car") # why doesn't it work?
summary(fit1)

```


##  Nested logit

One way to test the assumption is to estimate without one alternative and see if it affects the parameters. For instance we can focus on whether `air` or `train` is chosen and estimate within.

```{r}
fit.nested <- mlogit(choice ~ wait + gcost, TravelMode, reflevel = "car",
           nests = list(fly = "air", ground = c("train", "bus", "car")),
           unscaled = TRUE)

summary(fit.nested)
```

```{r}
data2 =copy(data)
I=paste(data2$mode)=="bus"

# force other alternatives to mean value
for (mm in c('car','train','air')) {
  #data2$gcost[paste(data2$mode)==mm] = mean(data2$gcost[paste(data2$mode)==mm])
  data2$wait[paste(data2$mode)==mm]  = mean(data2$wait[paste(data2$mode)==mm])
}


# run a for lopp for different prices
# save shares for each option
rr = foreach(dprice = seq(-100,100,l=20), .combine = rbind)  %do% {
  data2$gcost[I] = data$gcost[I] + dprice
  res = colMeans(predict(fit.nested,newdata=data2))
  res['dprice'] = dprice
  res
} 

rr = melt(data.frame(rr),id.vars = "dprice")
ggplot(rr,aes(x=dprice,y=value,color=factor(variable))) + geom_line()
```

## Random coefficient model

Let's try with 2 groups of people to run an EM

```{r}
C = acast(data,individual ~ mode,value.var="choice")
C = C[,c(4,1,2,3)]
p1=0.5

I = sample(unique(data$individua),nrow(data)/8)
I  =data$individual %in% I

# we start with the very first mlogit (we randomly sub-sample to create some variation)
fit1 <- mlogit(choice ~ gcost , data = data[I,], reflevel = "car")
fit2 <- mlogit(choice ~ gcost , data = data[!I,], reflevel = "car")

liks = rep(0,15)
for (i in 1:15) {
  # for each individual we compute the posterior probability given their data
  p1v = predict(fit1,newdata=data)
  p2v = predict(fit2,newdata=data)
  
  p1v = rowSums(p1v * C)*p1
  p2v = rowSums(p2v * C)*(1-p1)
  
  liks[i] = sum(log(p1v+p2v))
  #cat(sprintf("ll=%f\n",ll))
  
  p1v  = p1v/(p1v+p2v)
  p1v  = as.numeric( p1v %x% c(1,1,1,1) )
  
  # finally we run the 2 mlogit with weights
  fit1 <- mlogit(choice ~ gcost , data = data,weights = p1v, reflevel = "car")
  fit2 <- mlogit(choice ~ gcost , data = data,weights = as.numeric(1-p1v), reflevel = "car")
  
  p1 = mean(p1v)
}
print(fit1)
print(fit2)
print(p1)
```

```{r}

plot(liks)

```

